---
title: "rdantic"
format: 
  gfm:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
---

```{r}
#| include: false
source("rdantic.R")
```

**Declared types, checked at the boundary.** pydantic's idea, in one file of base R.

```r
source("rdantic.R")   # base R only; jsonlite is needed for the JSON helpers
```

| | |
|---|---|
| **Types are values** | `int[1]`, `num[. > 0]`, `chr[1] \| NULL` — built from `[` and `\|`, no new syntax |
| **Length is part of the type** | `int` is any integer vector, `int[1]` is exactly one |
| **Every problem, with a path** | not the first failure — all of them, each saying where |
| **Lossless coercion only** | `1` → `1L` yes; `1.5` → `1L` no; `"1"` → `1L` never |
| **Records and functions** | `model()` for validated objects, `fn()` for typed functions |
| **JSON in, JSON out** | `from_json()`, `to_json()`, `schema()` |

> **Status: proof of concept.** No S4, no R6, no code generation — closures, attributes
> and plain lists.
> **Feedback wanted!**


## Motivation

A payload arrives from somewhere you don't control. Plain R happily computes an answer.

```{r}
payload <- list(id = "1042", qty = 3, unit_price = 19.99, coupon = 1.2)

total <- function(o) o$qty * o$unit_price * (1 - o$coupon)
total(payload)
```

A negative total. No warning, no error. `id` is a string, `coupon` is above 1, and the
bug surfaces three calls later, in a report. Most R bugs look like this: not a crash, but
a value that was the wrong shape all along.

Declare what the data *is*, once:

```{r}
#| error: true
Order <- model("Order",
  id         = int[1][. > 0],
  qty        = int[1][. > 0],
  unit_price = num[1][. >= 0],
  coupon     = num[1][0 <= . & . <= 1] %default% 0
)

Order(id = "1042", qty = 3, unit_price = 19.99, coupon = 1.2)
```

Both mistakes at once, each naming the field and the rule it broke. Good data goes
through, coerced where that is lossless:

```{r}
order <- Order(id = 1042, qty = 3, unit_price = 19.99)
order
str(order$id)                            # the double 1042 arrived as an integer
```

Instances stay valid for their whole life, and a typed function can neither be called
nor return the wrong thing:

```{r}
#| error: true
order$qty <- 0

net <- fn(o = Order, ~ num[1][. >= 0], {
  o$qty * o$unit_price * (1 - o$coupon)
})
net(order)
net(payload)
```

That is the whole idea. The rest of this document is that idea applied to vectors,
records, nested data and JSON.

## Types

The four primitives cover R's atomic vectors; `[ ]` adds length and constraints, `|`
adds alternatives, and a handful of constructors add structure.

```{r}
int; num; chr; lgl
```

```{r}
int[1]
num[. > 0]
chr[1] | NULL
no_na(num)
one_of("low", "mid", "high")
fct("low", "mid", "high")
list_of(int[1])
list_of(cpu = int[1], memory = int[1])
frame(id = int, label = chr)
```

Because types are values, you name them and reuse them:

```{r}
id_t    <- int[1][. > 0]
email_t <- chr[1][grepl("@", ., fixed = TRUE)]
```

A type is callable: calling it validates a value and hands it back, coerced when that
loses nothing.

```{r}
#| error: true
int(c(1, 2, 3))          # whole doubles  -> integer
num(2L)                  # integer        -> double
chr(factor("a"))         # factor         -> character
int(1.5)                 # lossy
int("3")                 # parsing strings is never implicit
lgl(1)
```

Every error names *what was expected* and *what arrived* — type, length, and for scalars
the value. That shape is the same everywhere in the system.

Want no coercion at all? Values must then already be the right R type:

```{r}
#| error: true
options(rdantic.strict = TRUE)
int(1)
options(rdantic.strict = FALSE)
```

## Length

R's central data structure is the vector, so a type without a length is a type of
vectors. `T[n]` fixes the length; `T[1]` is a scalar.

```{r}
#| error: true
int[3](c(1, 2, 3))
int[3](c(1, 2))
int[1](c(1, 2))
```

The `[1]` matters more than it looks: most bugs from R's recycling rules come from a
vector arriving where one value was meant.

## Constraints

Anything inside `[ ]` that mentions `.` is a predicate over the value. It must return
`TRUE` for every element. Constraints and lengths chain in any order.

```{r}
#| error: true
num[. > 0](c(1, 2))
num[. > 0](c(1, -2))

int[1][. > 5](9)
int[1][. > 5](3)

chr[1][nchar(.) <= 3]("abcd")
```

The constraint's source text becomes part of the type's name, so the error tells the
reader the rule that was broken, not just that one was.

## Unions, optionals, defaults

`A | B` accepts either. `T | NULL` is the idiom for optional. `%default%` attaches a
default value, used by models and typed functions when a value is not supplied.

```{r}
#| error: true
(int[1] | chr[1])(7)
(int[1] | chr[1])("seven")
(int[1] | chr[1])(TRUE)

(int[1] | NULL)(NULL)

one_of("admin", "user") %default% "user"
int[1] | NULL %default% NULL              # a default on either side applies to the union
```

## Dates, factors, missing values

R's other quiet failure modes: a `Date` is a number wearing a class, a factor is an
integer wearing labels, and `NA` sails through every operator. Each gets a type.

```{r}
#| error: true
date("2024-05-17")                        # ISO-8601 is unambiguous, so it is parsed
date("17/05/2024")
num(Sys.Date())                           # ... but a Date is never silently a number
fct("low", "high")("high")
no_na(num)(c(1, NA))
```

When nothing built in fits, `type_from()` takes a plain predicate and an optional
coercion:

```{r}
#| error: true
hex <- type_from("hex", function(x) is.character(x) && all(grepl("^#[0-9a-f]{6}$", x)))
hex("#00ff99")
hex("green")
```

## Named lists

Give `list_of()` one type and you get a map: any keys, one value type. Keys are kept, and
a failure is reported by key rather than by position — this is what a JSON object with
arbitrary keys parses into.

```{r}
#| error: true
budget <- list_of(cpu = int[1][. > 0], memory = int[1][. > 0])

budget(list(cpu = 4, memory = 16))
budget(list(cpu = 4, memory = 0, disk = "big"))
```

Name the arguments instead and you get the opposite: a fixed key set with one rule per
key — still a plain named list on the way out, with no class and no model name.

```{r}
#| error: true
limits <- list_of(cpu = int[1][0 < . & . < 12], memory = int[1][0 < . & . < 128])
limits

str(limits(list(cpu = 4, memory = 16)))
limits(list(cpu = 16, memory = 512))
limits(list(cpu = 4))
to_json(limits(list(cpu = 4L, memory = 16L)))
```

Missing keys are reported, `%default%` fills them in, and `.extra = "forbid"` refuses
unknown ones — the same record rules `model()` uses. Reach for a model when the record
deserves a name, a class and inheritance; reach for `list_of(a = T, ...)` when it is just
a shape.

## Records: `model()`

A model is a set of named, typed fields. `model()` returns a constructor, and that
constructor is itself a type (so models nest and can be used anywhere a type is).

```{r}
User <- model("User",
  id     = id_t,
  name   = chr[1],
  email  = email_t,
  age    = int[1] | NULL,
  role   = one_of("admin", "user") %default% "user",
  tags   = chr %default% character(),
  friend = opt("User")                    # optional, refers to this model by name
)
User
```

Construct one like any R function — the constructor has real formals, so autocomplete
works:

```{r}
ada <- User(id = 1, name = "Ada", email = "ada@lovelace.org")
ada
str(ada$id)                              # coerced to integer on the way in
```

A field is required unless it has a default or accepts `NULL`. Every problem is reported:

```{r}
#| error: true
User(name = 42, email = "nope", role = "god")
```

Fields are validated on assignment too, and the object is sealed — a typo cannot quietly
invent a field, or quietly read a different one:

```{r}
#| error: true
ada$age <- 36L
ada$age
ada$age <- "thirty-six"
ada$nickname <- "Countess"
ada$nam
```

Instances are plain lists, so they behave like every other R value: copy on modify, no
action at a distance.

```{r}
copy <- ada
copy$name <- "Grace"
c(ada$name, copy$name)
identical(ada, User(id = 1, name = "Ada", email = "ada@lovelace.org", age = 36L))
```

## Nesting, and errors that report everything

Since a model is a type, models compose. Where a model is expected, a plain list is
accepted and parsed — this is what makes JSON input work later.

```{r}
Team <- model("Team",
  lead    = User,
  members = list_of(User),
  roster  = frame(id = int, name = chr)
)

bob  <- User(id = 2, name = "Bob", email = "bob@example.org", friend = ada)
team <- Team(lead = ada, members = list(bob),
             roster = data.frame(id = c(1, 2), name = c("Ada", "Bob")))
team
team$members[[1]]$friend$name
```

Three mistakes at three depths, one error:

```{r}
#| error: true
Team(
  lead    = list(id = 1, name = "Ada", email = "ada.lovelace"),
  members = list(list(id = 2, name = "Bob", email = "b@b", age = 34.5)),
  roster  = data.frame(id = c(1, 2.5), name = c("a", "b"))
)
```

Errors are machine-readable: a condition of class `typed_error` whose `$problems` is a
list of `path` / `expected` / `got` / `hint` records, so tooling can consume it, not just
people.

```{r}
tryCatch(User(id = "x", name = 1), typed_error = function(e)
  data.frame(
    path     = vapply(e$problems, `[[`, "", "path"),
    expected = vapply(e$problems, `[[`, "", "expected"),
    got      = vapply(e$problems, `[[`, "", "got")
  ))
```

When wrong input is expected rather than exceptional — a web form, an upload — use
`try_parse()`, which returns the same records instead of throwing:

```{r}
r <- try_parse(User, list(id = 1.5, name = "Ada", email = "nope"))
r$ok
str(r$problems[[2]])
is_valid(User, list(id = 1, name = "Ada", email = "a@b"))
```

## Inheritance, partials, unknown fields

```{r}
Admin <- extend(User, "Admin", permissions = list_of(chr[1]))
root  <- Admin(id = 1, name = "root", email = "root@example.org",
               permissions = list("all"))
inherits(root, "User")
ada$friend <- root                       # an Admin is a User

UserPatch <- partial(User)               # every field optional, defaults dropped
UserPatch(name = "Grace")
```

When parsing from a list, unknown keys are ignored by default (as in pydantic, and so
that JSON written from a subclass still parses as the parent). A model can forbid them:

```{r}
#| error: true
Strict <- model("Strict", x = int[1], .extra = "forbid")
from_list(Strict, list(x = 1, y = 2))
```

## Typed functions: `fn()`

Types sit where R would put defaults; an unnamed formula `~ T` declares the return type;
the last unnamed expression is the body. Defaults use `%default%`; arguments whose type
accepts `NULL` may be omitted.

```{r}
#| error: true
greet <- fn(name = chr[1], greeting = chr[1] %default% "Hello", ~ chr[1], {
  paste0(greeting, ", ", name, "!")
})
greet
greet("Ada")
greet("Ada", "Bonjour")
greet(c("Ada", "Bob"))
```

The return value is checked as well — a typed function cannot leak a wrong result:

```{r}
#| error: true
half <- fn(x = int[1], ~ int[1], { x / 2 })
half(4)                                  # 2 is a whole double: coerces to 2L
half(3)                                  # 1.5 is not an int
```

Give `...` a type to check everything passed through it. Every bad argument is reported
in one error, as with models:

```{r}
#| error: true
log_line <- fn(level = one_of("info", "warn"), ... = chr[1], ~ chr[1], {
  paste0("[", level, "] ", paste(..., collapse = " "))
})
log_line("info", "disk", "full")
log_line("debug", "disk", 3)
```

Checks are compiled once, when the function is defined, and cost tens of microseconds per
call — nothing at an API boundary, noticeable in a tight loop. Turn them off globally
when you no longer want to pay for them:

```{r}
options(rdantic.check = FALSE)
half(3)
options(rdantic.check = TRUE)
```

## Composing typed functions

Since a type is just a value plus a predicate, a *function* type needs no new machinery:

```{r}
pricing_rule <- anything[inherits(., "typed_fn")]
Money        <- num[1][. >= 0]
```

`rule_for()` **returns** a typed function; `invoice()` **takes** one. Three typed
functions, stacked:

```{r}
rule_for <- fn(kind = one_of("standard", "bulk", "clearance"), ~ pricing_rule, {
  switch(kind,
    standard  = fn(o = Order, ~ Money, { o$qty * o$unit_price }),
    bulk      = fn(o = Order, ~ Money, { o$qty * o$unit_price * 0.9 }),
    clearance = fn(o = Order, ~ Money, { o$qty * o$unit_price - 500 })
  )
})

invoice <- fn(o = Order, rule = pricing_rule, ~ Money, {
  round(rule(o), 2)
})

invoice(order, rule_for("bulk"))
```

The `clearance` rule is well-typed on the way in and wrong on the way out. Nothing
catches it until the arithmetic has already run:

```{r}
#| error: true
invoice(order, rule_for("clearance"))
```

The report names the frame that produced the bad value (`rule()`, the innermost of the
three) and the stage (`<return>`), not the outer call the user made. Compare the other
three places the same pipeline can break — each one is caught as early as it possibly
can be:

```{r}
#| error: true
rule_for("half off")                     # (1) argument, at the door
invoice(order, function(o) 0)            # (2) argument: a function, but not a typed one
invoice(payload, rule_for("bulk"))       # (2) argument: raw list, parsed then rejected
```

And a factory that forgets to return a function is caught by its own return type,
before anyone can call the result:

```{r}
#| error: true
broken_for <- fn(kind = chr[1], ~ pricing_rule, { toupper(kind) })
broken_for("bulk")
```

## JSON and schema

`to_json()` is type-aware: a field declared `[1]` becomes a JSON scalar, any other
vector becomes an array, even when it has one element.

```{r}
grace <- User(id = 3, name = "Grace", email = "grace@navy.mil", tags = "cobol")
to_json(grace)
```

`from_json()` parses and validates in one step. JSON arrays come back as R vectors and
arrays of objects as data frames, because the types accept those shapes — there is no
separate JSON mode.

```{r}
#| error: true
team2 <- from_json(Team, to_json(team))
str(team2$roster)
from_json(User, '{"id": 1.5, "name": ["a", "b"], "email": "x@y"}')
```

And a JSON Schema for interoperating with anything else that speaks types:

```{r}
to_json(schema(User))
```

## Cheat sheet

| Want | Write |
|---|---|
| integer / double / string / logical vector | `int`, `num`, `chr`, `lgl` |
| exactly n elements | `T[n]` |
| constraint | `T[. > 0]`, `chr[1][nchar(.) < 20]` |
| optional | `T \| NULL`, `opt(T)`, `opt("ModelName")` |
| either | `A \| B` |
| enum | `one_of("a", "b")` |
| date / time / factor | `date`, `datetime`, `fct("a", "b")` |
| no missing values | `no_na(T)` |
| custom type | `type_from("name", predicate, coerce)` |
| list of | `list_of(T)` |
| named list with fixed keys | `list_of(a = T, b = T)` |
| data.frame with typed columns | `frame(col = T, ...)` |
| a function value | `anything[is.function(.)]`, `anything[inherits(., "typed_fn")]` |
| default | `T %default% value` |
| record | `model("Name", f = T, ...)` |
| forward / self reference | `ref("Name")`, `opt("Name")` |
| subclass, partial | `extend(M, "Sub", ...)`, `partial(M)` |
| reject unknown keys | `model(..., .extra = "forbid")`, `frame(..., .extra = "forbid")` |
| typed function | `fn(a = T, ..., ~ Ret, { body })` |
| typed `...` | `fn(a = T, ... = T, ~ Ret, { body })` |
| validate anything | `T(x)`, `parse_as(T, x)`, `from_list(M, x)` |
| validate without throwing | `try_parse(T, x)`, `is_valid(T, x)` |
| fields of a model or value | `fields(M)`, `fields(x)` |
| no coercion / no checks | `options(rdantic.strict = TRUE)`, `options(rdantic.check = FALSE)` |
| JSON | `to_json(x)`, `from_json(T, txt)`, `schema(T)` |
| catch errors | `tryCatch(..., typed_error = function(e) e$problems)` |

## How it works

* **Types are closures** carrying a `spec` attribute (`name`, `validate`, `schema`).
  `[.type` captures its argument with `substitute()` — a bare number is a length, an
  expression mentioning `.` is a predicate. `Ops.type` implements `|`. That is the whole
  DSL; there is no parser.
* **Models compile to constructors** with real formals, built with `formals<-` and
  `bquote()`. Instances are **plain lists** carrying their own `spec`, so they copy on
  modify like every other R value and `identical()`, `saveRDS()` and `str()` behave;
  `$<-` dispatches to a method that validates and returns the updated value.
* **Validators return the value itself**, or a `typed_problems` list of records. The
  happy path allocates nothing beyond the value, and `rdantic.strict` is read once per
  top-level validation rather than once per element.
* **`fn()` rewrites the function**: types come off the formals, one check block is
  prepended, the return is wrapped. R 4.4 reserved `declare()` for exactly this kind of
  annotation; when it gains semantics, `fn()` can emit it and the checks move into the
  interpreter without user code changing.
* **Coercion is lossless only**: whole doubles → int (never out of range), int → double,
  factor → character, ISO-8601 text → date. While parsing JSON, also: unnamed list of
  scalars → vector, list of records → data frame. Nothing else, ever.

## Not in this proof of concept

* Discriminated unions for polymorphic JSON (`{"type": "admin", ...}` choosing the model).
* Generic models (`Page(of = User)`).
* Cross-field validators (`.check = function(self) ...`) and row-wise constraints in
  `frame()` beyond predicates over a column.
* Packaging all of this as an actual R package, with a namespace instead of `source()`.
