---
title: "Containers: lists, maps and frames"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Containers: lists, maps and frames}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rdantic)
```

R has three shapes for "more than one of something", and rdantic has one container type
for each: `list_of()` for an array, `map_of()` for arbitrary-key/one-value-type maps, and
`frame()` for a data.frame with typed columns. `list_of()` also has a second form for a
fixed key set. This vignette covers when to reach for which; see `vignette("structs")` for
when a record deserves a name and a class instead.

## `list_of(T)`: an array of one type

`list_of(T)` accepts an unnamed list of any length, all elements of type `T`. A named
list is rejected -- that is `map_of()`'s job, so the two never overlap.

```{r, error = TRUE}
tags <- list_of(chr[1])
tags(list("a", "b"))
tags(list(a = "x"))              # a named list is map_of(), not this
```

## `map_of(T)`: arbitrary keys, one value type

`map_of(T)` is the map counterpart: any keys, one value type, kept by name. A failure is
reported by key rather than by position -- and this is what a JSON object with arbitrary
keys parses into (see `vignette("json-and-schemas")`).

```{r, error = TRUE}
budget <- map_of(int[1][. > 0])
budget(list(cpu = 4, memory = 16))
budget(list(cpu = 4, memory = 0, disk = "big"))
map_of(int[1])(list(1, 2))       # unnamed input is list_of(), not this

schema(budget)
```

## `list_of(a = T, b = T, ...)`: a fixed key set

Name the arguments of `list_of()` instead of giving it one type, and it becomes a third
shape: a fixed set of keys with one rule per key -- still a plain named list on the way
out, with no class and no struct name.

```{r, error = TRUE}
limits <- list_of(cpu = int[1][0 < . & . < 12], memory = int[1][0 < . & . < 128])
limits

str(limits(list(cpu = 4, memory = 16)))
limits(list(cpu = 16, memory = 512))
limits(list(cpu = 4))                    # missing key, reported
```

Missing keys are reported, `%default%` fills them in, and `.extra = "forbid"` refuses
unknown ones -- the same record rules `struct()` uses:

```{r, error = TRUE}
list_of(cpu = int[1], .extra = "forbid")(list(cpu = 1, junk = 2))
```

Reach for a struct (`vignette("structs")`) when the record deserves a name, a class and
inheritance; reach for `list_of(a = T, ...)` when it is just a shape.

## `frame()`: a data.frame with typed columns

`frame()` describes a data.frame: each column is validated as a vector, so the usual
length and constraint syntax applies to it. A column of the wrong length for the number
of rows, or a missing column, is an error.

```{r, error = TRUE}
roster <- frame(id = int, name = chr)
str(roster(data.frame(id = c(1, 2), name = c("Ada", "Bob"))))

roster(data.frame(id = 1))               # missing column
frame(id = int, .extra = "forbid")(data.frame(id = 1L, junk = 2))
```

While parsing JSON, an array of objects is reshaped into a data frame, and a row missing
a key is reported by name instead of quietly producing a list column:

```{r, error = TRUE}
from_json(roster, '[{"id": 1, "name": "Ada"}, {"id": 2, "name": "Bob"}]')
from_json(roster, '[{"id": 1}]')
nrow(from_json(roster, "[]"))
```

## Picking one

| Shape of the data | Use |
|---|---|
| unnamed, any length, one element type | `list_of(T)` |
| named, arbitrary keys, one value type | `map_of(T)` |
| named, fixed keys, one rule each, no class needed | `list_of(a = T, b = T, ...)` |
| rows and columns | `frame(col = T, ...)` |
| a record that deserves a name, a class, inheritance | `struct("Name", field = T, ...)` |
