---
title: "Error handling and validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Error handling and validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rdantic)
```

Every validation failure in rdantic -- from a bare type, a struct, or a typed function --
has the same shape: a list of problems, each naming the path where it occurred, what was
expected, what arrived, and (often) a hint about why. This vignette covers the three ways
to get at that information, and when to reach for each.

## Errors are machine-readable

By default, a failed validation throws a condition of class `typed_error` whose
`$problems` field is a list of `path` / `expected` / `got` / `hint` records -- so tooling
can consume it, not just people.

```{r}
User <- struct("User", id = int[1], name = chr[1])

tryCatch(User(id = "x", name = 1), typed_error = function(e)
  data.frame(
    path     = vapply(e$problems, `[[`, "", "path"),
    expected = vapply(e$problems, `[[`, "", "expected"),
    got      = vapply(e$problems, `[[`, "", "got")
  ))
```

Because `typed_error` is a distinct condition class, `tryCatch()` can single it out from
other errors a call might raise.

## Validating without throwing

When wrong input is *expected* rather than exceptional -- a web form, a file upload, a
config file -- throwing on every bad value is the wrong shape. `try_parse()` returns the
same problem records instead of raising them:

```{r}
r <- try_parse(User, list(id = 1.5, name = "Ada"))
r$ok
str(r$problems[[1]])
```

`try_parse()` returns a list with `ok`, `value` (the validated/coerced value when `ok` is
`TRUE`) and `problems` (always a list, empty when there is nothing to report).

## A yes/no answer

When only "is this valid" matters, not why, `is_valid()` skips building any problem detail
and returns a plain `TRUE`/`FALSE`:

```{r}
is_valid(User, list(id = 1, name = "Ada"))
is_valid(User, list(id = 1.5, name = "Ada"))
```

## Picking one

| Situation | Use |
|---|---|
| Wrong input is a bug in the caller | let it throw, or `tryCatch(..., typed_error = ...)` |
| Wrong input is routine (forms, uploads, config) | `try_parse(T, x)` |
| Only need a yes/no answer | `is_valid(T, x)` |
| The type is in a variable, not a literal | `parse_as(T, x)` (same as `T(x)`) |

## Turning checks off

Validation has a real, if small, cost. `fn()` checks can be disabled globally with
`options(rdantic.check = FALSE)` once a typed function's contract is trusted -- see
`vignette("typed-functions")`. There is no equivalent switch for structs or bare types:
their whole purpose is that an instance can never become invalid, so disabling their
checks is not offered.
