---
title: "Extending rdantic: new_type()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending rdantic: new_type()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rdantic)
```

Most custom rules fit inside `type_from()` (`vignette("types-in-depth")`): a predicate,
an optional coercion, and rdantic looks after the problem format for you. This vignette
covers `new_type()`, the lower-level constructor that everything else in the package --
`int`, `list_of()`, `struct()`, `fn()` -- is itself built from, for when a type needs
full control over its validation and its JSON Schema fragment.

## What a type actually is

A type is a closure that validates a value, carrying a `spec` attribute describing it
(`name`, `validate`, `schema`, and whatever else the constructor stored). Calling a type
runs `validate()`; printing it, or asking `schema()` for its JSON Schema, reads the rest
of the spec. `new_type()` is how that closure gets built:

```{r}
even <- new_type(
  "even",
  validate = function(x, path) {
    if (is.numeric(x) && all(x %% 2 == 0)) x else rdantic:::.bad(path, "even", type_of(x))
  },
  schema = function() list(type = "integer", multipleOf = 2)
)
even
even(c(2, 4))
```

```{r, error = TRUE}
even(3)
```

`validate` receives the value and the `path` at which it appears (so an error nested
inside a struct or a list reports the right location) and returns either the -- possibly
coerced -- value, or a problem record built by rdantic's internal helpers. `type_of()`
(exported, used above) renders a value the way rdantic names types in its own error
messages, so a custom type's errors read the same as a built-in one's.

```{r}
schema(even)
```

## When to reach for this vs. `type_from()`

| Need | Use |
|---|---|
| A predicate, maybe a coercion, default JSON Schema shape | `type_from()` |
| Full control over the problem message, or a custom JSON Schema fragment | `new_type()` |
| A restricted set of exact values | `one_of()` |
| A record with named fields | `struct()` |

In practice `new_type()` is rarely needed directly -- it exists so the rest of rdantic can
be implemented in terms of it, and remains available for the rare case a project's own
domain type needs the same level of control.
