---
title: "JSON and schemas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{JSON and schemas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rdantic)   # jsonlite must also be installed for the JSON helpers
```

A struct declared once is also a JSON contract: `schema()` builds a JSON Schema fragment
from the same declaration that `to_json()` and `from_json()` use to serialize and
validate, so the two can never drift apart. This is especially useful for talking to a
structured-output API: send the schema as the response format, then run whatever comes
back through `from_json()` before trusting it.

## Serializing

`to_json()` is type-aware: a field declared `[1]` becomes a JSON scalar, any other vector
-- even one with a single element -- becomes a JSON array.

```{r}
User <- struct("User",
  id   = int[1][. > 0],
  name = chr[1],
  role = one_of("admin", "user") %default% "user",
  tags = chr %default% character()
)

grace <- User(id = 3, name = "Grace", tags = "cobol")
to_json(grace)
```

## Parsing and validating

`from_json()` parses and validates in one step. A JSON array of scalars comes back as an
R vector, and an array of objects becomes a data frame, because the declared types accept
those shapes -- there is no separate "JSON mode".

```{r, error = TRUE}
from_json(User, '{"id": 1, "name": "Ada"}')
from_json(User, '{"id": -1, "name": "Ada"}')   # validated against the same declaration
from_json(User, '{"id": 1.5, "name": ["a", "b"]}')
```

Round-tripping a struct through JSON returns an equivalent instance:

```{r}
grace2 <- from_json(User, to_json(grace))
identical(grace$name, grace2$name)
```

## Schemas

`schema()` builds a JSON Schema list; pass it to `to_json()` to serialize it for sending
to an API.

```{r}
to_json(schema(User))
```

What `schema()` declares always matches what `to_json()` actually writes. A `frame()`
column's schema describes one row's scalar cell, not the whole column's vector shape, and
`map_of()`'s schema is a JSON object, matching the object it serializes to (see
`vignette("containers")` for both):

```{r}
Cell <- struct("Cell", rows = frame(id = int), tags = map_of(chr[1]))
c(schema(Cell)$properties$rows$items$properties$id$type,
  schema(Cell)$properties$tags$type)
to_json(Cell(rows = data.frame(id = 1L), tags = list(a = "x")), pretty = FALSE)
```

## Documenting a schema

A description belongs in the schema itself, not beside it in a comment that can drift out
of sync -- this is what a structured-output model actually reads. There are two places to
attach one: `desc()` (or its infix spelling `%doc%`) documents a single field's type, and
`struct(.description = )` documents the record as a whole.

```{r}
Contact <- struct("Contact",
  .description = "A single contact record.",
  name  = chr[1] %doc% "Full name of the person.",
  email = opt(chr[1]) %doc% "Email address, or null if unknown."
)

to_json(schema(Contact))
```

`desc()`/`%doc%` wraps a type, so -- like `%default%` -- it must be the outermost part of
a chain, applied after `[n]`, `[expr]`, `opt()`/`|` and `%default%`:

```{r}
age <- int[1][. > 0] %doc% "Age in years, must be positive."
schema(age)
```

## An end-to-end example

Putting it together: declare the shape once, hand the schema to a model, and validate
whatever it returns.

```{r, error = TRUE}
Extraction <- struct("Extraction",
  name = chr[1],
  age  = int[1][. > 0],
  tags = list_of(chr[1])
)

to_json(schema(Extraction))   # send this as the response format

reply <- '{"name": "Ada", "age": -1, "tags": ["r", "stats"]}'
from_json(Extraction, reply)  # a reply that ignores the schema is still caught
```
