---
title: "S7 compatibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{S7 compatibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("S7", quietly = TRUE)
)
```

```{r setup}
library(rdantic)
library(S7)
```

[S7](https://rconsortium.github.io/S7/) is the R Consortium's new object system: classes
with named, typed properties, single and multiple dispatch, and `@` for property access.
rdantic and S7 answer different halves of the same question -- S7 says what a class *is*,
rdantic says what a value must *satisfy* and reports, with a path, every place the
incoming data failed to. They compose in both directions, and neither takes the other
over: S7 is a suggested dependency, so nothing in this vignette exists unless you have
S7 installed.

## S7 in one minute

If S7 is new to you, four ideas are enough for this vignette:

* `new_class()` creates a **class**. Calling that class creates an **object**.
* A class declares named **properties**, each with an S7 class such as
  `class_double`. Properties are supplied to the constructor by name.
* `x@lat` reads a property and `x@lat <- value` sets it. S7 checks the property's
  class when the object is constructed or changed.
* S7 generics dispatch on the class of one or more arguments, much like S3 methods but
  with an explicit class definition.

For example, `Pin` below is a class, `Pin(lat = 1, lon = 2)` is an object, and `lat`
and `lon` are properties. This is only a working introduction; the
[S7 documentation](https://rconsortium.github.io/S7/) covers classes, properties and
dispatch in depth.

## Choose a direction

| Starting point | Goal | Use |
|----------------|------|-----|
| An existing S7 class | Parse, validate or describe incoming data | Use the class directly, or call `s7_type()` |
| A new model described with rdantic types | Produce S7 objects | `s7_struct()` |
| An existing rdantic struct | Hand it to S7-based code | `as_s7_class()` |

The rest of the vignette follows those three paths in that order.

## Using an S7 class as a type

An S7 class can be written anywhere an rdantic type is expected. `as_type()` -- which
`struct()`, `fn()`, `list_of()` and friends all call -- recognises it and hands it to
`s7_type()`:

```{r}
Pin <- new_class("Pin", properties = list(
  lat = class_double,
  lon = class_double
))

as_type(Pin)
```

That makes it a field type like any other. A plain list is parsed into a real S7 object,
which is what makes JSON work with no separate JSON mode:

```{r}
Trip <- struct("Trip",
  label = chr[1],
  start = Pin,
  stop  = Pin
)

t <- from_json(Trip, '{
  "label": "commute",
  "start": {"lat": 52.52, "lon": 13.40},
  "stop":  {"lat": 48.14, "lon": 11.58}
}')

t
class(t$start)
t$start@lat
```

Problems inside a property carry the path they occurred at, and every problem is
reported at once:

```{r, error = TRUE}
from_list(Trip, list(
  label = "commute",
  start = list(lat = "north", lon = 13.40),
  stop  = list(lat = 48.14)
))
```

An S7 object that is already of the right class is passed through untouched; one of a
different class is rejected rather than silently re-parsed:

```{r, error = TRUE}
Trip(label = "ok", start = Pin(lat = 1, lon = 2), stop = 42)
```

Schemas, `to_list()` and `to_json()` all understand S7 objects too:

```{r}
schema(Pin)
to_json(t$start)
```

### What carries over, and what does not

An S7 property declares a class; an rdantic type declares a class *and a shape*. The
translation keeps what transfers and refuses to invent the rest:

| S7 property class                                             | rdantic type              |
|---------------------------------------------------------------|---------------------------|
| `class_integer`, `class_double`, `class_character`, `class_logical` | `int`, `num`, `chr`, `lgl` |
| `class_list`                                                   | `list_of(anything)`       |
| `class_Date`, `class_POSIXct`                                  | `date`, `datetime`          |
| `class_data.frame`                                             | an unconstrained data frame |
| a `new_union()`                                                | `A \| B`                  |
| another S7 class                                               | that class, recursively   |
| anything else, including `class_any`                           | `anything`                |

Note that `class_double` becomes `num`, not `num[1]`: S7 places no length constraint on a
property, so neither does rdantic, and `to_json()` therefore writes a JSON array. A
property is **required** unless it declares a `default`, which is the same rule
`%default%` follows for struct fields.

```{r}
Vector <- new_class("Vector", properties = list(x = class_double))
from_list(Vector, list(x = c(1, 2, 3)))@x
```

Native S7's `new_class()` does not accept rdantic expressions such as `num[1]` as
property classes. To require exactly one double, define the S7 class through rdantic:

```{r, error = TRUE}
Scalar <- s7_struct("Scalar", x = num[1])
Scalar(x = 1)
Scalar(x = c(1, 2))
```

`class_data.frame` similarly says only that the value is a data frame; it says nothing
about column names or types. rdantic preserves that meaning. Use an rdantic
`frame(id = int, name = chr)` field when you want a particular table shape.

### S7 details and boundaries

The conversion keeps S7's own behavior around the translated fields:

* A property with a getter but no setter is computed. It is omitted from parsing,
  schemas and serialization because callers cannot supply it. A property with both a
  getter and setter remains a field.
* Constant and expression-valued S7 defaults make a property optional. Expressions are
  left for S7 to evaluate for each new object rather than being evaluated by rdantic.
* The class's S7 validator still runs when rdantic constructs an object. An abstract S7
  class can describe and validate existing objects, but a list cannot be parsed into a
  new instance of it.
* Unknown input keys are ignored when parsing into an existing S7 class. Classes made
  by `s7_struct()` can instead use `.extra = "forbid"`.

An existing object of the requested class is returned unchanged. Use `from_list()` or
`from_json()` on incoming data when parsing and path-aware validation at a boundary is
the goal.

## Declaring an S7 class with rdantic types

`s7_struct()` is `struct()` with an S7 class as the result. The fields are written
exactly as they would be for a struct, and every property validates through its rdantic
type -- on construction *and* on `@<-`:

```{r}
Account <- s7_struct("Account",
  id     = int[1][. > 0],
  email  = chr[1][grepl("@", ., fixed = TRUE)],
  credit = num[1][. >= 0] %default% 0
)

a <- Account(id = 1, email = "ada@example.org")
a@credit
```

Coercion happens exactly where rdantic would coerce it, so the property holds the value
rdantic promised rather than the one you happened to type:

```{r}
a@credit <- 10L
typeof(a@credit)
```

and the failures are the same `typed_error`s, with the property named as `@field`:

```{r, error = TRUE}
a@id <- 0
```

```{r, error = TRUE}
Account(id = 1)
```

The result is an ordinary S7 class, so S7 generics dispatch on it:

```{r}
balance <- new_generic("balance", "x")
method(balance, Account) <- function(x) x@credit
balance(a)
```

and an ordinary rdantic type, so it parses, validates and describes itself:

```{r}
from_json(Account, '{"id": 2, "email": "bob@example.org"}')
schema(Account)$properties$id
```

Properties are declared `class_any`, deliberately: the real constraint is the rdantic
type, and a second, weaker check by S7 would only be able to disagree with it. The field
types are kept on the class, which is why `int[1][. > 0]` survives above rather than
decaying to `anything`.

One deliberate difference from `struct()`: the name is **not** registered, so `ref()`
and `opt("Account")` will not find it. A name in that registry means something that
builds instances, and this builds S7 objects. Pass the class itself instead -- it is a
value, and a type.

### Inheritance

`.parent` takes an S7 class, whose properties are inherited as typed fields:

```{r}
Premium <- s7_struct("Premium", tier = one_of("gold", "silver"), .parent = Account)

p <- Premium(id = 1, email = "ada@example.org", tier = "gold")
class(p)
fields(Premium)
```

There is one case rdantic can express and S7 cannot. rdantic lets a child **retype** an
inherited field; S7 keeps the parent's property definition and silently ignores a
redeclaration. Rather than hand back a class whose narrower type is never enforced, the
hierarchy is flattened -- the type is honoured, the `S7_inherits()` relationship is not
-- and you are told:

```{r}
Big <- s7_struct("Big", id = int[1][. > 100], .parent = Account)
class(Big(id = 101, email = "ada@example.org"))
```

```{r, error = TRUE}
Big(id = 1, email = "ada@example.org")
```

## Converting an existing struct

When the struct already exists -- because other code parses into it, or it is reached by
`ref()` -- `as_s7_class()` converts it rather than redeclaring it:

```{r}
Animal <- struct("Animal", name = chr[1], legs = int[1])
Dog    <- extend(Animal, "Dog", good = lgl[1] %default% TRUE)

d <- as_s7_class(Dog)(name = "Rex", legs = 4L)
class(d)
S7_inherits(d, as_s7_class(Animal))
```

`extend()` becomes S7 inheritance, under the same flattening rule as above. Unlike
`s7_struct()`, `as_s7_class()` caches by name and field signature, so converting the same
struct twice returns the same class and `S7_inherits()` keeps working.

`extend()` and `partial()` also accept an S7 class directly, and read its properties as
typed fields -- so you can go back to plain lists at any point:

```{r}
fields(extend(Account, "AccountNote", note = chr[1]))
fields(partial(Account))
```

## In short

Use an S7 class directly when it is already your domain model and rdantic belongs at
the input boundary. Use `s7_struct()` when you are declaring the model with rdantic
types but downstream code expects S7. Use `as_s7_class()` when the rdantic struct
already exists and you need an S7 view of it.
