---
title: "Structs: modeling records"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structs: modeling records}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rdantic)
```

A struct is a set of named, typed fields. `struct()` returns a constructor, and that
constructor is itself a type -- so structs nest inside other structs, functions and
containers, and can appear anywhere a type is expected. This vignette assumes the basics
from `vignette("types-in-depth")`.

## Defining and constructing

```{r}
User <- struct("User",
  id     = int[1][. > 0],
  name   = chr[1],
  email  = chr[1][grepl("@", ., fixed = TRUE)],
  age    = int[1] | NULL,
  role   = one_of("admin", "user") %default% "user",
  tags   = chr %default% character()
)
User
```

Construct one like any R function -- the constructor has real formals, so autocomplete
works in an IDE:

```{r}
ada <- User(id = 1, name = "Ada", email = "ada@lovelace.org")
ada
str(ada$id)                              # coerced to integer on the way in
```

A field is required unless it has a `%default%` or its type accepts `NULL`. Every problem
in a bad call is reported at once, not just the first:

```{r, error = TRUE}
User(name = 42, email = "nope", role = "god")
```

## Instances stay valid

Fields are validated on assignment as well as construction, and a struct is sealed: a
typo cannot quietly invent a new field, or quietly read a different one. `$<-`, `[[<-`
and `[<-` all run the same check, so there is no assignment form that skips it:

```{r, error = TRUE}
ada$age <- 36L
ada$age
ada$age <- "thirty-six"
ada["age"] <- "thirty-six"               # [<- checks the field just like $<- does
ada$nickname <- "Countess"
ada$nam
```

Instances are otherwise plain lists (with a class), so they behave like every other R
value: copy on modify, no action at a distance.

```{r}
copy <- ada
copy$name <- "Grace"
c(ada$name, copy$name)
```

A field name that collides with rdantic's own internal bookkeeping (like `.args`) is
rejected when the struct is declared, rather than silently shadowed:

```{r, error = TRUE}
struct("Broken", .args = int[1])
```

## Nesting

Where a struct is expected, a plain list is accepted and parsed into it -- this is what
lets JSON input work later (see `vignette("json-and-schemas")`) with no separate JSON
mode.

```{r, error = TRUE}
Team <- struct("Team",
  lead    = User,
  members = list_of(User)
)

bob  <- User(id = 2, name = "Bob", email = "bob@example.org")
team <- Team(lead = ada, members = list(bob))
team$members[[1]]$name

# three mistakes at three depths, one error
Team(
  lead    = list(id = 1, name = "Ada", email = "ada.lovelace"),
  members = list(list(id = 2, name = "Bob", email = "b@b", age = 34.5))
)
```

## Self- and forward-reference

A struct that needs to refer to its own type (or one defined later) cannot write its own
name as a bare symbol -- `ref("Name")` resolves the name lazily, on first use, and caches
it; `opt("Name")` is the common `ref("Name") | NULL` shorthand for an optional link:

```{r}
Node <- struct("Node", value = int[1], next_ = opt("Node"))
Node(value = 1, next_ = list(value = 2))
```

## Inheritance and partials

`extend()` subclasses a struct: the child gets the parent's fields plus its own, inherits
the parent's class (so it is accepted anywhere the parent is), and may retype a parent
field by repeating its name.

```{r}
Animal <- struct("Animal", name = chr[1], legs = int[1] %default% 4L)
Bird   <- extend(Animal, "Bird", can_fly = lgl[1])

tweety <- Bird(name = "Tweety", legs = 2, can_fly = TRUE)
inherits(tweety, "Animal")

# a child may retype a parent field
fields(extend(Animal, "Centipede", legs = int[1][. > 50]))
```

`partial()` makes every field of a struct optional -- for PATCH-shaped input, where only
the supplied keys mean anything. Defaults are dropped along with requirements, so "not
supplied" stays distinguishable from "set to the default":

```{r}
Post <- struct("Post", title = chr[1], body = chr[1], draft = lgl[1] %default% TRUE)
PostPatch <- partial(Post)

patch <- PostPatch(title = "New title")
patch$title
patch$draft          # NULL, not the default
```

## Unknown fields

When parsing from a list, unknown keys are ignored by default -- as in pydantic, and so
that JSON written from a subclass still parses as the parent. Pass `.extra = "forbid"` to
reject them instead:

```{r, error = TRUE}
Strict <- struct("Strict", x = int[1], .extra = "forbid")
from_list(Strict, list(x = 1, y = 2))
```

## Introspection and plain-list conversion

`fields()` lists a struct's (or an instance's) field names in declaration order.
`to_list()` recursively turns an instance -- and any nested instances -- back into plain
named lists, e.g. for code that should not know about rdantic:

```{r}
fields(User)
fields(ada)

Pin  <- struct("Pin", lat = num[1], lon = num[1])
Trip <- struct("Trip", from = Pin, to = Pin)
str(to_list(Trip(from = Pin(lat = 1, lon = 2), to = Pin(lat = 3, lon = 4))))
```

See `vignette("json-and-schemas")` for `to_json()`/`from_json()`/`schema()`, and
`vignette("errors-and-validation")` for reading and handling the errors a struct raises.
