---
title: "Dates and temporal types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dates and temporal types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rdantic)
```

R has three quiet failure modes this vignette's types guard against: a `Date` is a
number wearing a class, a factor is an integer wearing labels, and a plain string is
neither until something parses it. rdantic gives each of these its own type instead of
letting them silently interoperate with plain numbers and strings.

## Dates and datetimes

`date` accepts `Date` vectors, `datetime` accepts `POSIXct` vectors. Unlike plain numbers,
ISO-8601 text *is* parsed -- it is the one unambiguous serialization of an instant -- but
never silently to `NA`: text that does not parse is an error, not a missing value.

```{r, error = TRUE}
date("2024-05-17")                        # ISO-8601 is unambiguous, so it is parsed
date(as.Date("2024-05-17"))
datetime("2024-05-17T09:30:00")

date("17/05/2024")                        # wrong layout, not a silent NA
date(19860)                               # a number is not a date
num(Sys.Date())                           # ... and a date is never silently a number
```

Like any other type, a date can carry a length or be used inside a union or a container:

```{r}
date[1]                                   # exactly one date
parse_as(date, c("2024-01-01", "2024-06-30"))
```

## Restricted factors

`fct(a, b, ...)` accepts a factor with exactly those levels, and promotes character input
whose values are all among them. Use `one_of()` (`vignette("types-in-depth")`) when a
plain string should come out; use `fct()` when a factor with a fixed, ordered level set
should.

```{r, error = TRUE}
size <- fct("low", "mid", "high")
size("mid")                               # character, promoted to a factor
levels(size(c("high", "low")))
size("enormous")                          # not one of the declared levels

fct()(factor("anything"))                 # fct() with no levels: any factor, no coercion
```

## Missing values

`NA` sails through most R operators without complaint. `no_na(T)` wraps any type to
additionally reject it -- see `vignette("types-in-depth")` for `no_na()` alongside the
other combinators.

```{r, error = TRUE}
no_na(date)(as.Date(c("2024-01-01", NA)))
```
