---
title: "Types in depth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Types in depth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rdantic)
```

A type in rdantic is an ordinary R value: a closure you can call, print, store in a
variable, and pass to `struct()` or `fn()`. This vignette covers the vocabulary for
building one out of R's own atomic vectors, `[ ]` and `|`. For records and functions see
`vignette("structs")` and `vignette("typed-functions")`; for lists, maps and data frames
see `vignette("containers")`.

## The four primitives

`int`, `num`, `chr` and `lgl` cover R's atomic vector types. Each one, called on a value,
validates it and hands it back -- coerced only when that coercion cannot lose anything.

```{r}
int; num; chr; lgl
```

```{r, error = TRUE}
int(c(1, 2, 3))          # whole doubles  -> integer
num(2L)                  # integer        -> double
chr(factor("a"))         # factor         -> character
int(1.5)                 # lossy: rejected
int("3")                 # parsing strings is never implicit
lgl(1)                   # 0/1 is not TRUE/FALSE
```

Every error names what was expected and what arrived -- type, length, and for scalars the
value itself. That shape is the same everywhere in rdantic, whether the failure came from
a bare type, a struct field or a function argument.

Because types are values, name them and reuse them across a codebase:

```{r}
id_t    <- int[1][. > 0]
email_t <- chr[1][grepl("@", ., fixed = TRUE)]
```

Use them exactly like any other type -- call them to validate:

```{r, error = TRUE}
id_t(1)
id_t(c(0, 1))

email_t("something@example.com")
email_t("This is not an email")
```

### Strict mode

By default, rdantic coerces where nothing is lost. Turn that off to require the exact R
type:

```{r, error = TRUE}
options(rdantic.strict = TRUE)
int(1)                    # a double, even a whole one, no longer coerces
options(rdantic.strict = FALSE)
```

## Length

R's central data structure is the vector, so a type without a length is a type of
vectors. `T[n]` fixes the length to exactly `n`; `T[1]` is the common scalar case.

```{r, error = TRUE}
int[3](c(1, 2, 3))
int[3](c(1, 2))
int[1](c(1, 2))
```

`[1]` matters more than it looks: most bugs from R's recycling rules come from a vector of
the wrong length arriving where a single value was meant.

For a range rather than an exact length, use `length(.)`: inside `[ ]`, `.` is bound to
the whole vector, not one element at a time, so `length(.)` is the vector's length.

```{r, error = TRUE}
int[length(.) <= 3](c(1, 2))
int[length(.) <= 3](c(1, 2, 3, 4))

int[length(.) >= 10](1:10)
int[length(.) >= 10](1:5)
```

## Constraints

Anything inside `[ ]` that mentions `.` is a predicate over the value, applied
elementwise. Length and constraints chain in any order, and can be combined:

```{r, error = TRUE}
num[. > 0](c(1, 2))
num[. > 0](c(1, -2))

int[1][. > 5](9)
int[1][. > 5](3)

chr[1][nchar(.) <= 3]("abcd")
```

The constraint's source text becomes part of the type's name, so the error message tells
the reader the rule that was broken, not just that some rule was.

## Unions, optionals and defaults

`A | B` accepts either type. `T | NULL` (or the shorthand `opt(T)`) is how to say
"optional". `%default%` attaches a value used when nothing is supplied -- by a struct
field or a function argument, see `vignette("structs")` and `vignette("typed-functions")`.

```{r, error = TRUE}
(int[1] | chr[1])(7)
(int[1] | chr[1])("seven")
(int[1] | chr[1])(TRUE)

(int[1] | NULL)(NULL)
opt(int[1])(NULL)

one_of("admin", "user") %default% "user"
```

## Enums, missing values and custom rules

`one_of(...)` restricts a value to an exact set of options. `no_na(T)` wraps any type to
additionally reject `NA`.

```{r, error = TRUE}
one_of("low", "mid", "high")("high")
one_of("low", "mid", "high")("medium")

no_na(num)(c(1, 2))
no_na(num)(c(1, NA))
```

For dates, times and restricted factors, see `vignette("temporal-types")`.

When nothing built in fits, `type_from()` builds a type from a plain predicate and an
optional coercion, without touching rdantic's internal problem format:

```{r, error = TRUE}
hex <- type_from("hex", function(x) is.character(x) && all(grepl("^#[0-9a-f]{6}$", x)))
hex("#00ff99")
hex("green")

port <- type_from(
  "port",
  function(x) is.integer(x) && length(x) == 1 && x > 0 && x < 65536,
  coerce = function(x) as.integer(x),
  json = "integer",
  scalar = TRUE
)
port(8080)
```

For a type that needs full control over its validation logic and JSON Schema fragment,
see `new_type()` in `vignette("extending-rdantic")`.

## Validating a value directly

A type is callable, but three other entry points exist for when the type is not known
ahead of time, or when a thrown error is not what you want -- see
`vignette("errors-and-validation")` for the full picture:

```{r, error = TRUE}
parse_as(int[1], 3)             # same as int[1](3), useful when the type is a variable
is_valid(int[1], 3.5)           # TRUE/FALSE, no detail
try_parse(int[1], "x")$ok       # validate without throwing
```
