---
title: "Typed functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Typed functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rdantic)
```

`fn()` declares a function whose arguments and return value are both checked. Types sit
where R would put defaults; an unnamed formula `~ T` declares the return type; the last
unnamed expression is the body.

## Basics

```{r, error = TRUE}
greet <- fn(name = chr[1], greeting = chr[1] %default% "Hello", ~ chr[1], {
  paste0(greeting, ", ", name, "!")
})
greet
greet("Ada")
greet("Ada", "Bonjour")
greet(c("Ada", "Bob"))          # a vector where one value was meant
```

`%default%` supplies a value used when an argument is omitted, exactly as it does for a
struct field (`vignette("structs")`); an argument whose type accepts `NULL` may also be
omitted.

## The return value is checked too

A typed function cannot leak a wrong result. An explicit `return()` goes through exactly
the same check as a value that falls through at the end of the body:

```{r, error = TRUE}
half <- fn(x = int[1], ~ int[1], { x / 2 })
half(4)                                  # 2 is a whole double: coerces to 2L
half(3)                                  # 1.5 is not an int

early <- fn(x = int[1], ~ int[1], { if (x < 0) return("negative"); x })
early(-1)                                # return()'s value is checked too
```

## Reserved names and typed `...`

An argument name that collides with rdantic's own internal bookkeeping (like `.probs_`)
is rejected when the function is declared, rather than silently shadowed:

```{r, error = TRUE}
fn(.probs_ = int[1], ~ int[1], { .probs_ })
```

Give `...` a type to check every value passed through it. Values passed via `...` are
checked but not coerced (`...` cannot be rebound), and every bad argument is reported in
one error, as with a struct:

```{r, error = TRUE}
log_line <- fn(level = one_of("info", "warn"), ... = chr[1], ~ chr[1], {
  paste0("[", level, "] ", paste(..., collapse = " "))
})
log_line("info", "disk", "full")
log_line("debug", "disk", 3)
```

## Turning checks off

Checks are compiled once, when the function is defined, and cost tens of microseconds per
call -- nothing at an API boundary, but possibly noticeable in a tight inner loop. Turn
them off globally when that cost matters more than the safety:

```{r}
options(rdantic.check = FALSE)
half(3)                                  # no longer checked
options(rdantic.check = TRUE)
```

## Composing typed functions

`typed_fn` accepts any function created by `fn()`. More often, higher-order code needs a
particular signature: `fn_type(o = Order, ~ Money)` accepts only a typed function with
that declared argument and return type.

```{r}
Order <- struct("Order",
  id         = int[1][. > 0],
  qty        = int[1][. > 0],
  unit_price = num[1][. >= 0],
  coupon     = num[1][0 <= . & . <= 1] %default% 0
)
order <- Order(id = 1042, qty = 3, unit_price = 19.99)

Money        <- num[1][. >= 0]
pricing_rule <- fn_type(o = Order, ~ Money)
```

`rule_for()` **returns** a typed function; `invoice()` **takes** one. Three typed
functions, stacked:

```{r}
rule_for <- fn(kind = one_of("standard", "bulk", "clearance"), ~ pricing_rule, {
  switch(kind,
    standard  = fn(o = Order, ~ Money, { o$qty * o$unit_price }),
    bulk      = fn(o = Order, ~ Money, { o$qty * o$unit_price * 0.9 }),
    clearance = fn(o = Order, ~ Money, { o$qty * o$unit_price - 500 })
  )
})

invoice <- fn(o = Order, rule = pricing_rule, ~ Money, {
  round(rule(o), 2)
})

invoice(order, rule_for("bulk"))
```

Signature checks include argument names and defaults. They compare declarations rather
than calling the function, so no user code runs during validation. Use the broader
`typed_fn` type when any declared signature is acceptable.

The `clearance` rule is well-typed on the way in and wrong on the way out (a negative
total). Nothing catches it until the arithmetic has already run:

```{r, error = TRUE}
invoice(order, rule_for("clearance"))
```

The report names the frame that produced the bad value (`rule()`, the innermost of the
three) and the stage (`<return>`), not the outer call the user made. Compare the other
places the same pipeline can break -- each is caught as early as it possibly can be:

```{r, error = TRUE}
rule_for("half off")                     # (1) argument, at the door
invoice(order, function(o) 0)            # (2) argument: a function, but not a typed one
```

And a factory that forgets to return a function is caught by its own return type, before
anyone can call the result:

```{r, error = TRUE}
broken_for <- fn(kind = chr[1], ~ pricing_rule, { toupper(kind) })
broken_for("bulk")
```
